---
title: "Medical Malpractice"
output: html_document
author: "Elsie Yi"
date: "2023-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Nearly [50% of physicians age 55 and older have reported being sued](https://www.ama-assn.org/practice-management/sustainability/1-3-physicians-has-been-sued-age-55-1-2-hit-suit) according to the American Medical Association and in [2022, 31% of physicians have been sued](https://www.ama-assn.org/practice-management/sustainability/medical-liability-market-research). 
With the prevalence of these malpractice suits I was interested to see if there is any association between the amount received by the patient and whether they had a private attorney, what type of insurance they had (private, uninsured, or government), severity of damage, as well as age and gender of patient. 
I am also curious to see if there is a greater proportion of suits given the type of specialty of the physician.

Load packages
```{r}
library(ggplot2)
library(GGally)
```

Load data
```{r}
mm <- read.csv("/Users/elsieyi/Documents/Projects/Kaggle/Medical Malpractice/medicalmalpractice.csv")
mm$severe_range <- cut(mm$Severity, breaks = c(min(mm$Severity),3,6,max(mm$Severity)), include.lowest= TRUE)
mm$age_range <- cut(mm$Age, breaks = c(min(mm$Age), 18, 25, 35, 45, 55, 65, max(mm$Age)), include.lowest=TRUE)
```

## Dataset

Data is from [Kaggle](https://www.kaggle.com/datasets/gabrielsantello/medical-malpractice-insurance-dataset/). The data set contains information about 79,210 claim payments made. 

Visualization was completed with Tableau and R and the analysis is completed in R. 


### Initial data insights

#### Data variables
* **Amount**: claim payment in dollars
* **Severity**: rating of damage to patient. 1 (emotional trauma) to 9 (death) || adjusted into range later on
* **Age**: age of patient in years
* **Private Attorney**: 0: not represented by private attorney || 1: represented by private attorney
* **Specialty**: physician specialty 
* **Insurance**: patient's medical insurance
* **Gender**: patient's gender

Lowest amount of malpractice suit payment:
```{r}
min(mm$Amount)
```

Highest amount of malpractice suit payment:
```{r}
max(mm$Amount)
```

## Exploratory Data Analysis 
As severity of damage to the patient increases, so does the amount of claim payment.
```{r}
as_set <- aggregate(Amount ~ severe_range, mm, mean)
ggplot(data=as_set, aes(x = severe_range, y=Amount)) +
  geom_bar(stat="identity", width=0.5) +
  scale_y_continuous(labels= scales::comma) + ggtitle("Average Claim Payment Amount vs Severity of Damage")
```

Proportion of those in specified age groups in the damage severity groups. A majority of those in the [18,25], [25,35], [55,65], and [65,87] age group experienced damage in the [1,3] severity level whereas a majority in the [0,18], (35,45], and (45,55] age groups had severity level range of (3,6]
```{r}
prop.table(table(mm$age_range, mm$severe_range))
```


## Modeling
#### Private Attorney
Without consideration for other variables, it can be seen that having a private attorney is statisitcally significant and is expected to increase the amount one can receive in the malpractice suit by an estimated \$106,848. 
```{r}
pa <- lm(Amount ~ Private.Attorney, data = mm)
summary(pa)
```

#### Insurance Type
There is evidence that insurance type is statistically significant for the amount of payment for all types: No Insurance, Private Insurance, Medicare/Medicaid, Workers Compensation, and Unknown.
Private insurance is expected to increase the amount of payment compared to a patient who does not have insurance, however, one can expect to receive a reduction in payment given medicare/medicaid, workers compensation, or unknown insurance. 
```{r}
mm$Insurance <- relevel(factor(mm$Insurance), ref = "No Insurance")
ins <- lm(Amount ~ Insurance, data = mm)
summary(ins)
```

#### Severity
Adjusting severity into 3 ranges: less severe (1-3), severe (4-6), most severe (7-9).
As a patient's severity of trauma 
```{r}
sev <- lm(Amount ~ severe_range, data = mm)
summary(sev)

sev2 <- lm(Amount ~ severe_range + Age, data = mm)
summary(sev2)
```

#### Specialty
I chose the comparison group to be 'Pediatrics' as according to [past studies](https://www.ama-assn.org/system/files/policy-research-perspective-medical-liability-claim-frequency.pdf) this is a specialty in which fewer physicians faced lawsuits. 
```{r}
sp <- lm(Amount ~ Specialty, data = mm)
summary(sp)
```


#### Age and Gender
There is evidence suggesting that age and gender are statistically significant in amount of malpractice suit payment but not for an interaction term between them. This suggests that the effect age has on the amount received is not affected by gender. A 1 year increase in age is expected to result in an estimated reduction of \$1,196.53 in amount received and males are expected to receive an estimated \$2,2638.15 more.

A reduction in expected claim payment as age increases could be due to a greater presence of comorbidities seen in older patients. 
```{r}
ag <- lm (Amount ~ Age + Gender + Age*Gender, data = mm)
summary(ag)
```

#### All variables
```{r}
all.model <- lm(Amount ~ Private.Attorney + severe_range + Age + Gender + Insurance + Specialty, data = mm)
summary(all.model)
```

